---
title: "PSYC2020L Assignment 8"
author: "[name]"
date: "2025-10-10"
output: html_document
---

## Goals for Review:

* Importing
* NHST


## Setting Up

### 1) Rename this assignment as "Lab 8 Assignment [Last, First].Rmd" 

### 2) Check and set (if needed) your working directory. If you'd like to load packages with `library()`, please do so in this block.
```{r}

```


## Import

### 3) Download the `testdata.csv` file from Canvas, and relocate it to the folder for this lab. Import the csv into R. Try knitting the document right after this question for simpler troubleshooting! This data contains three variables:

| name  | description         | notes                             |
|-------|---------------------|-----------------------------------|
| `id`  | examinee identifier |                                   |
| `SAT` | SAT score           | can theoretically range 400--1600 |
| `ACT` | ACT score           | can theoretically range 1--36     |

Display the first few rows of this data. 
```{r}

```


## EDA

### 4) Explore or report on the SAT variable in three different ways. Examples include making a plot, commenting on the shape of the distribution, numerical summary statistics (e.g, mean, SD, range), etc.
```{r}

```

### 4) Explore or report on the ACT variable in three different ways. Examples include making a plot, commenting on the shape of the distribution, numerical summary statistics (e.g, mean, SD, range), etc.
```{r}

```


## *Z*-Tests

You've been tasked with testing whether the scores from this sample of students differ from the population means.

### 5a) Let's say the population mean for SAT scores is 1200 and the population SD is 110. What are your null and alternative hypotheses?



### 5b) Calculate the *z*-test statistic for this sample of SAT scores against the population.
```{r}

```

### 5c) What is the *p*-value of that *z*-statistic?
```{r}

```

### 5d) Provide an interpretation for this *z*-test given the test statistic and *p*-value.



### 6a) Let's say the population mean for ACT scores is 29 and the population variance is 64. What are your null and alternative hypotheses?



### 6b) Calculate the *z*-test statistic for this sample of ACT scores against the population.
```{r}

```

### 6c) What is the *p*-value of that *z*-statistic?
```{r}

```

### 6d) Provide an interpretation for this *z*-test given the test statistic and *p*-value.








