---
title: "PSYC2020L Assignment 4"
author: "[name]"
date: "2025-09-12"
output: html_document
---

## Goals for Review:

* Probability Distributions Review
* *Z*-Scores in R
* Generating Data
* R Packages
* Reading Files

For each question, please make sure you include code output. If the question asks for comparison or further explanation, please make sure to include that as well. 


## Setting Up

### 1) Rename this assignment as "Lab 4 Assignment [Last, First].Rmd" 

### 2) Check and set (if needed) your working directory. If you'd like to load packages with `library()`, please do so in this block.
```{r}

```

### 3) Read in the `tour_de_france.csv` file and assign it to a variable (e.g., `tdf_data`). Display the first six rows of this dataset.
```{r}

```


## Measures of Variability

### 4) What is the variance and SD of `winner_avg_speed`?
```{r}

```

### 5) What is the variance and SD of `total_distance`?
```{r}
```


## Z-Scores

### 6) Create a new variable within the Tour de France data that contains the `winner_avg_speed` variable standardized. No output needed for this question.
```{r}
```

### 7) Create a new variable within the Tour de France data that contains the `total_distance` variable standardized. No output needed for this question.
```{r}
```

### 8) Calculate the variance and SD of the new *z*-scored version `winner_avg_speed`, and compare it to the original, non-*z*-scored version. What differences do you observe? 
```{r}

```


## Generation from Normal Distribution

### 9) Generate three vectors and assign them to variables: one with 10 draws from a standard normal distribution, one with 50 draws from a standard normal distribution, and one with 100 draws from a standard normal distribution. Display just the vector with 10 draws.
```{r}

```

### 10) Plot a histogram of the vector out of the three that will appear the most representative of the underlying standard normal distribution.
```{r}

```

### 11) Create two more vectors and assign them to variables: one with 1000 draws from a normal distribution with mean of zero and SD of 2 and one with 1000 draws from a normal distribution with mean of 2 and variance of 4. No output is needed for this question.
```{r}

```

### 12) Plot each of the two vectors from (11) in a histogram. Set the x-axis limits to be the same for both (hint: use the ? tool if you don't remember how to set axis limits!), and compare the two plots. 
```{r}

```


## Matrices

### 13) Create a vector of sequential integers from 1 to 10 and assign it to a variable. No output is needed for this question.
```{r}

```

### 14) Turn this vector in a 2 rows x 5 columns matrix that fills in the rows first. Assign this matrix to a variable and display the matrix as output.
```{r}

```


## Indexing

### 15) From that matrix, please index:

#### 15a) The second row.
```{r}

```

#### 15b) The third column of the first row.
```{r}

```

#### 15c) The second through fourth columns of the first and second rows.
```{r}

```











