---
title: "PSYC2020L Assignment 9"
author: "[name]"
date: "2025-10-17"
output: html_document
---

## Goals for Review:

* Confidence Intervals
* One Sample T-Tests


## Setting Up

### 1) Rename this assignment as "Lab 9 Assignment [Last, First].Rmd" 

### 2) Check and set (if needed) your working directory. If you'd like to load packages with `library()`, please do so in this block. 
```{r}
library(tidyverse)

here::here()
```


## Confidence Intervals

### 3a) Calculate the 95% confidence interval around the mean for iris sepal length, and report its interpretation.

```{r}
Rmisc::CI(iris$Sepal.Length)
```

If we were to repeat this data collection many times, 95% of the time, the true mean will fall within the interval [`r Rmisc::CI(iris$Sepal.Length)[3]`, `r Rmisc::CI(iris$Sepal.Length)[1]`].

### 3b) Would you expect the 99% confidence interval around the mean for iris sepal length to be wider or narrower than that of the 95th? What about the 67% confidence interval?

The 99th would be wider and the 67th would be narrower.

### 3c) Calculate the 67% confidence interval around the mean for iris sepal length, and report its interpretation.

```{r}
Rmisc::CI(iris$Sepal.Length, ci = .67)
```

If we were to repeat this data collection many times, 95% of the time, the true mean will fall within the interval [`r Rmisc::CI(iris$Sepal.Length, ci = .67)[3]`, `r Rmisc::CI(iris$Sepal.Length, ci = .67)[1]`].


## One-Sample *T*-Test

### 4a) A professor claims that students spend 4.5 hours per day studying. You're not so sure! You are given a dataset (`studytimes.csv`) that contains the study hours of 25 students. Your goal is to load the dataset into R and analyze whether the mean study time is significantly different from 4.5 hours using a one-sample *t*-test. First, load the data in and display the first few rows.

```{r}
study <- rio::import(here::here("labs", "Lab 9 - Confidence Intervals", "studytimes.csv"))
head(study)
mean(study$study_time)
sd(study$study_time)
```

### 4b) Write the null and alternative hypotheses for a two-tailed test.

Null: The sample mean of study time is equal to 4.5 hours.

Alternative: The sample mean of study time is not equal to 4.5 hours.

### 4c) Perform a one-sample *t*-test using the dataset.

```{r}
t.test(study$study_time, mu = 4.5)
```

### 4d) Report the *p*-value and interpret the results at α = 0.05.

The *p*-value is .384. We found no significant difference between the sample mean and the hypothesized value of 4.5 hours.

### 4e) Report the 95% confidence interval for the mean study time. Does the confidence interval include 4.5 hours? What does that mean in terms of hypothesis testing?

The 95% CI is [4.30, 5.01]. Because this contains our hypothesized value of 4.5 hours, we know the difference between the sample mean and hypothesized mean is not significant at the α = .05 level.

### 4f) What is the effect size of the difference between the sample mean and the hypothesized mean?

```{r}
(mean(study$study_time) - 4.5) / sd(study$study_time)
```

### 4g) Interpret the effect size.

The Cohen's *d* value was 0.178. This is a small effect.
