---
title: "PSYC2020L Assignment 9"
author: "[name]"
date: "2025-10-17"
output: html_document
---

## Goals for Review:

* Confidence Intervals
* One Sample T-Tests
* Effect Size


## Setting Up

### 1) Rename this assignment as "Lab 9 Assignment [Last, First].Rmd" 

### 2) Check and set (if needed) your working directory. If you'd like to load packages with `library()`, please do so in this block. 
```{r}

```


## Confidence Intervals

### 3a) Calculate the 95% confidence interval around the mean for iris sepal length, and report its interpretation.

```{r}

```



### 3b) Would you expect the 99% confidence interval around the mean for iris sepal length to be wider or narrower than that of the 95th? What about the 67% confidence interval?

.

### 3c) Calculate the 67% confidence interval around the mean for iris sepal length, and report its interpretation.

```{r}

```




## One-Sample *T*-Test

### 4a) A professor claims that students spend 4.5 hours per day studying. You're not so sure! You are given a dataset (`studytimes.csv`) that contains the study hours of 25 students. Your goal is to load the dataset into R and analyze whether the mean study time is significantly different from 4.5 hours using a one-sample *t*-test. First, load the data in and display the first few rows.

```{r}

```

### 4b) Write the null and alternative hypotheses for a two-tailed test.



### 4c) Perform a one-sample *t*-test using the dataset.

```{r}

```

### 4d) Report the *p*-value and interpret the results at α = 0.05.



### 4e) Report the 95% confidence interval for the mean study time. Does the confidence interval include 4.5 hours? What does that mean in terms of hypothesis testing?



### 4f) What is the effect size of the difference between the sample mean and the hypothesized mean?

```{r}

```

### 4g) Interpret the effect size.


