---
title: "PSYC2020L Assignment 5"
author: "[name]"
date: "2025-09-19"
output: html_document
---

## Goals for Review:

* `tidyverse` concepts
* Import
* Tidy
* Transform
* Model
* Visualize
* Communicate

For each question, please make sure you include code output. If the question asks for comparison or further explanation, please make sure to include that as well. 


## Setting Up

### 1) Rename this assignment as "Lab 5 Assignment [Last, First].Rmd" 

### 2) Check and set (if needed) your working directory. If you'd like to load packages with `library()`, please do so in this block. Make sure to load the `tidyverse` package! Also, please install (feel free to delete the install code after installing) and load the `palmerpenguins` package.
```{r}

```


## Import

### 3) Now that we've loaded the `palmerpenguins` package, we have access to a datset called `penguins`. We also have access to a dataset called `penguins_raw`, but we will not use this one. `penguins` is already an object / variable we can use, but if you'd like to assign it to a different variable name (totally up to you), feel free to do that here. Either way, use three functions to get a look at this data.
```{r}

```

### 4) As may be obvious from the name, this dataset contains penguin measurements for different species! Use `?penguins` to read a little bit more about the variables present.
```{r}
#?penguins
```



## Tidy

### 5) Per the rules we learned in lab, is this a *tidy* dataset? How do you know?



## Transform & Model

Overview: as the statistician on this penguin project, you have been given two tasks:

1: Identify the mean and SD for flipper length by species

2: Identify the minimum and maximum bill length-to-depth ratio for female penguins on the Biscoe island by year.

3: Lastly, they want you to come up with and answer a question of your own with these data!

### 6a) Let's start with 1: Identify the mean and SD for flipper length by species. Starting with the `penguins` dataset, `summarize()` the data to get the mean and SD for `flipper_length_mm` by species. Don't forget about the `.by` argument for the `summarize()` function! 

*Note*: there will be missing values for this dataset. Add the `na.rm = TRUE` argument to any `mean()` or `SD()` function to remove those.
```{r}

```


### 6b) Discuss these findings (1--2 sentences). 



### 7a) Now let's move on to 2: Identify the minimum and maximum bill length-to-depth ratio for female penguins on the Biscoe island by year. In one pipe, do the following things:

* Calculate the bill length-to-depth ratio by adding a column that divides `bill_length_mm` by `bill_depth_mm`.

* Narrow down the data to the desired penguins by filtering such that only female penguins on the Biscoe island remain.

* Summarize the data with two values: the minimum bill length-to-depth ratio and the maximum bill length-to-depth ratio. Do this by the `year` the measurement was collected.
```{r}

```

### 7b) Discuss these findings (1--2 sentences). 


### 8a) Come up with your own question to answer about this dataset! As info, this question will need to include at least two functions' worth of `tidyverse` piping (i.e., data |> function1() |> function2()). First, just write your question below.

How does the mean of `body_mass_g` for female penguins vary over species and island?

### 8b) Next, write the code that addresses your question.
```{r}

```

### 8c) Lastly, please discuss these findings (1--2 sentences). 



## Visualize

Next, we will make two visualizations from this data. We will do this step by step like in the slides. 

Our first goal will be to make a scatterplot of bill length and depth with the dot color showing species. We will also add trend lines and some customization.

### 9a) To start, using the `ggplot()` function, set up a grid that uses the penguins data. Put `bill_length_mm` on the x-axis and `bill_depth_mm` on the y-axis.
```{r}

```

### 9b) Next, let's add some points. Feel free to copy down your code from the last question, and add the `geom_()` function that will create a scatter plot.
```{r}

```

### 9c) Please describe / discuss what you see in the plot above (1--2 sentences). Also---if R gave a warning, please describe what it's telling you! No need to make any changes from the warning.



### 9d) Now, let's let the color of the points represent species. First, do we need to modify the aesthetic or the geom? How do you know?



### 9e) Now, modify the plot such that point color represents species.
```{r}

```

### 9f) Please describe / discuss what you see in the plot above. How does the color indicator add information about the trends in these data? (1--2 sentences). 



### 9g) Next, we are going to add trend lines to this plot, also as by species. Do we need to modify the aesthetic or the geom, and how do you know?


### 9h) Add the trend lines! Include the argument `method = "lm"` in your change.
```{r}

```


### 9i) Lastly, please add some customization to this plot! Make at least three changes. For each change, please describe what you added / adjusted and why you think it betters the visualization.
```{r}

```



