---
title: "PSYC2020L Assignment 4"
author: "[name]"
date: "2025-09-12"
output: html_document
---

## Goals for Review:

* *Z*-Scores in R
* Generating Data
* R Packages
* Reading Files

For each question, please make sure you include code output. If the question asks for comparison or further explanation, please make sure to include that as well. 


## Setting Up

### 1) Rename this assignment as "Lab 4 Assignment [Last, First].Rmd" 

### 2) Check and set (if needed) your working directory. If you'd like to load packages with `library()`, please do so in this block.
```{r}
here::here()
```

### 3) Read in the `tour_de_france.csv` file and assign it to a variable (e.g., `tdf_data`). Display the first six rows of this dataset.
```{r}
tdf <- rio::import(here::here("labs", "Lab 4 - Z-Scores", "tour_de_france.csv"))
head(tdf)
```


## Measures of Variability

### 4) What is the variance and SD of `winner_avg_speed`?
```{r}
var(tdf$winner_avg_speed)
sd(tdf$winner_avg_speed)
```
<div style = "color:maroon">If they do some square or square root transformation to get one from the other, that works too. </div>

### 5) What is the variance and SD of `total_distance`?
```{r}
var(tdf$total_distance)
sd(tdf$total_distance)
```
<div style = "color:maroon">If they do some square or square root transformation to get one from the other, that works too. </div>


## Z-Scores

### 6) Create a new variable within the Tour de France data that contains the `winner_avg_speed` variable standardized. No output needed for this question.
```{r}
tdf$winner_avg_speed_z <- (tdf$winner_avg_speed - mean(tdf$winner_avg_speed)) / sd(tdf$winner_avg_speed)
```

### 7) Create a new variable within the Tour de France data that contains the `total_distance` variable standardized. No output needed for this question.
```{r}
tdf$total_distance_z <- (tdf$total_distance - mean(tdf$total_distance)) / sd(tdf$total_distance)
```

### 8) Calculate the variance and SD of the new *z*-scored version `winner_avg_speed`, and compare it to the original, non-*z*-scored version. What differences do you observe? 
```{r}
var(tdf$winner_avg_speed_z)
sd(tdf$winner_avg_speed_z)
```
The variance and SD for the z-scored version are both one. This is because the z-score transformation sets the variance <div style = "color:maroon">[or SD]</div> to one. This obviously much less than the original variance of 26.94 and SD of 5.19.


## Generation from Normal Distribution

### 9) Generate three vectors and assign them to variables: one with 10 draws from a standard normal distribution, one with 50 draws from a standard normal distribution, and one with 100 draws from a standard normal distribution. Display just the vector with 10 draws.
```{r}
norm10 <- rnorm(10, 0, 1)
norm50 <- rnorm(50, 0, 1)
norm100 <- rnorm(100, 0, 1)
norm10
```
### 10) Plot a histogram of the vector out of the three that will appear the most representative of the underlying standard normal distribution.
```{r}
hist(norm100)
```

### 11) Create two more vectors and assign them to variables: one with 1000 draws from a normal distribution with mean of zero and SD of 2 and one with 1000 draws from a normal distribution with mean of 2 and variance of 4. No output is needed for this question.
```{r}
norm02 <- rnorm(1000, 0, 2)
norm22 <- rnorm(1000, 2, sqrt(4))
```

### 12) Plot each of the two vectors from (11) in a histogram. Set the x-axis limits to be the same for both (hint: use the ? tool if you don't remember how to set axis limits!), and compare the two plots. 
```{r}
hist(norm02, xlim = c(-6, 6))
hist(norm22, xlim = c(-6, 6))
```
The histograms have the same variance, but the one with mean of 2 is shifted to the right.


## Matrices

### 13) Create a vector of sequential integers from 1 to 10 and assign it to a variable. No output is needed for this question.
```{r}
myvec <- 1:10
```

### 14) Turn this vector in a 2 rows x 5 columns matrix that fills in the rows first. Assign this matrix to a variable and display the matrix as output.
```{r}
mymat <- matrix(myvec, nrow = 2, ncol = 5, byrow = T)
mymat
```


## Indexing

### 15) From that matrix, please index:

#### 15a) The second row.
```{r}
mymat[2, ]
```
#### 15b) The third column of the first row.
```{r}
mymat[1, 3]
```
#### 15c) The second through fourth columns of the first and second rows.
```{r}
mymat[1:2, 2:4]
```











