---
title: "PSYC2020L Assignment 1"
author: "Jessica Helmer"
date: "2025-08-22"
output: html_document
---
<div style = "color:maroon">Take off a small number of points if no name. </div>
## Goals for Review:

* Basic R operators and data types
* Viewing data
* Basic plotting

## Setting Up

### 1) Rename this assignment as "Lab 1 Assignment [Last, First].Rmd" 

### 2) Check and set your working directory

```{r}
getwd()
```
<div style = "color:maroon">We ran out of time to cover setting working directory; this is all they have to do for this question.</div>

## R Basics

### 3) Compute:

#### a) The product of 4 * 4

```{r}
4 * 4
```

#### b) 4 squared

```{r}
4 ^ 2
# or
4 ** 2
```
<div style = "color:maroon">The first one is more likely; it's what I showed in class.</div>

#### c) The remainder from dividing 16 by 3

```{r}
16 %% 3
```

#### d) `FALSE` divided by `TRUE`. Why did you get this answer?

```{r}
FALSE / TRUE
```
`FALSE` and `TRUE` are represented as 0 and 1 by R. When you divide 0 by 1, you get 0.

### 4) The output of the following two lines of code is different. Why is that?

```{r}

3 + 2 / 2

```

```{r}

(3 + 2) / 2

```
R follows standard the order of operations for math (PEMDAS). Because parentheses are evaluated first, the (3 + 2) becomes 5 in the second version, which is then divided by 2 to get 2.5. In the first version, the 2 / 2 is evaluated first, which becomes 1 and is then added to 3 to get 4.

### 4) 

#### a) Create a vector of your four favorite movies
<div style="color:maroon">These don't have to have output as long as the syntax is correct. </div>
```{r}
my_movies <- c("Dune: Part 1", "Dune: Part 2", "Third Movie", "Fourth Movie")
```
<div style="color:maroon">They don't have to assign it to a variable here; creating a vector is enough. </div>

#### b) Create a vector of the years they came out

```{r}
movie_years <- c(2022, 2024, 2000, 2000)
```
<div style="color:maroon">They don't have to assign it to a variable here; creating a vector is enough. </div>

#### c) Create a vector of the ratings (number of stars out of five) that YOU think they deserve

```{r}
ratings <- c(5, 5, 5, 5)
```
<div style="color:maroon">They don't have to assign it to a variable here; creating a vector is enough. </div>

#### d) Put all those vectors in a dataframe with the column labels "fav_movie", "year", and "my_rating"
<div style="color:maroon">Same here, doesn't have to have output as long as the syntax is correct. </div>
```{r}
data.frame(fav_movie = my_movies,
           year = movie_years,
           my_rating = ratings)
```
<div style="color:maroon">Whether they retype it by hand or use variable assignment is fine as long as the responses from a, b, and c match d. The variable names should be `fav_movie`, `year`, and `my_rating`. </div>

### 5) Show me...

### The first three rows of the `iris` dataset

```{r}
head(iris, n = 3)
```

### 6) Plot...

#### A scatterplot with Petal.Length on the x and Petal.Width on the y

```{r}
plot(iris$Petal.Length, iris$Petal.Width)
```








