---
title: "PSYC2020L Assignment 1"
author: "[name]"
date: "2025-08-22"
output: html_document
---

## Goals for Review:

* Basic R operators and data types
* Viewing data
* Basic plotting

## Setting Up

### 1) Rename this assignment as "Lab 1 Assignment [Last, First].Rmd" 

### 2) Check and set your working directory

```{r}

```

## R Basics

### 3) Compute:

#### a) The product of 4 * 4

```{r}

```

#### b) 4 squared

```{r}

```

#### c) The remainder from dividing 16 by 3

```{r}

```

#### d) `FALSE` divided by `TRUE`. Why did you get this answer?

```{r}

```

### 4) The output of the following two lines of code is different. Why is that?

```{r}

3 + 2 / 2

```

```{r}

(3 + 2) / 2

```

### 4) 

#### a) Create a vector of your four favorite movies

```{r}

```

#### b) Create a vector of the years they came out

```{r}

```

#### c) Create a vector of the ratings (number of stars out of five) that YOU think they deserve

```{r}

```

#### d) Put all those vectors in a dataframe with the column labels "fav_movie", "year", and "my_rating"

```{r}

```

### 5) Show me...

### The first three rows of the `iris` dataset

```{r}

```

### 6) Plot...

#### A scatterplot with Petal.Length on the x and Petal.Width on the y

```{r}

```








