---
title: "Sampling Distributions"
subtitle: "PSYC 2020-A01 / PSYC 6022-A01 | 2025-09-26 | Lab 6"
author: "Jessica Helmer"
format: 
  live-revealjs:
    slide-number: true
    show-slide-number: all
    css: PSYC2020L_styles.css
engine: knitr
execute:
  echo: true
width: 1200
---

## Outline

* Assignment 5 Review
* Sampling Distributions

Learning objectives:  
**R:** `tidyverse`

```{r, echo =F}
library(tidyverse)
```


## Housekeeping

[input any housekeeping items here]


## Assignment 5 Review

[input any Assignment 5 review here]


## Sampling Distributions

Important topic in statistics! Help us make inferences about the estimates we calculate

:::: {.columns}
::: {.column}
#### Sample Distribution

Distribution of the data within our sample

:::

::: {.column}
#### Sample Distribution

(More theoretical) distribution of our statistic over repeated sampling
:::
::::


## Population

```{r}
sampcolor <- "turquoise"
popsize <- 1000
sampsize <- 50

population <- data.frame(id = 1:popsize,
                         x = rnorm(popsize)) %>%
  mutate(y = runif(popsize, min = 0, max = dnorm(x)))

sampling <- data.frame(sampmean = as.numeric())
         
population %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  coord_cartesian(xlim = c(-3, 3), y = c(0, .5), expand = F) +
  theme_classic(base_size = 14) +
  labs(y = NULL) +
  theme(axis.text.y = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        aspect.ratio = 1)
```

```{r}
sampdist <- function() {
  population <<- population %>%
    mutate(samp = factor(ifelse(x %in% sample(x, sampsize), 1, 0), ordered = T))
  
  population %>%
    ggplot(aes(x = x, y = y, color = samp)) +
    geom_point() +
    coord_cartesian(xlim = c(-3, 3), y = c(0, .5), expand = F) +
    theme_classic(base_size = 14) +
    scale_color_manual(NULL, values = c("black", sampcolor)) +
    labs(y = NULL) +
    guides(color = "none") +
    theme(axis.text.y = element_blank(),
          axis.line.y = element_blank(),
          axis.ticks.y = element_blank(),
          aspect.ratio = 1)
}

samplingdist <- function() {
  sampling <<- sampling %>%
    rbind(population %>% filter(samp == 1) %>% summarize(sampmean = sum(x) / sampsize))
  
  sampling %>%
    ggplot(aes(x = sampmean)) +
    geom_histogram(fill = sampcolor, color = NA, bins = 5, alpha = .6) +
    geom_density(color = sampcolor, linewidth = 1) +
    coord_cartesian(xlim = c(-3, 3), ylim = c(0, 5), expand = F) +
    theme_classic(base_size = 14) +
    labs(y = NULL) +
    guides(color = "none") +
    theme(axis.text.y = element_blank(),
          axis.line.y = element_blank(),
          axis.ticks.y = element_blank(),
          aspect.ratio = 1)
}
```


## Sample

:::: {.columns}
::: {.column}
```{r}
sampdist()
```
:::

::: {.column}
```{r}
samplingdist()
```
:::
::::




