---
title: "PSYC2020L Assignment 14"
author: "[name]"
date: "2025-11-21"
output: html_document
---

## Goals for Review:

* Correlation and Regression


## Setting Up

### 1) Rename this assignment as "Lab 14 Assignment [Last, First].Rmd" 

### 2) Check and set (if needed) your working directory. If you'd like to load packages with `library()`, please do so in this block.

```{r}
here::here()
```

### 3) Download the `cake_data.csv` dataset from Canvas. Relocate it to your folder for this class and import it into R. Please display the first few rows. Additionally, please use two other functions to get a sense of the data. This data contains the breakage angle `angle` for cakes made with different `recipes`. 

```{r}
cake <- rio::import(here::here("labs", "Lab 14 - One-Way ANOVA", "cake_data.csv"))
```


## Analysis of Variance (ANOVA)

### 4) You'd like to test to see if there are differences in the breakage angle for cakes made with these different recipes with a One-Way ANOVA. What are your null and alternative hypotheses for this test?

$H_0$: $\sigma^2_B = 0$

$H_1$: $\sigma^2_B \neq 0$

### 5) Perform this ANOVA and display the results with `summary()`.

```{r}
aov(angle ~ recipe, data = cake) |> summary()
```

### 6) Briefly describe your results.

There is a significant effect of recipe *F*(3, 96) = 39.1, *p* < .001 on cake breakage angle.

[more items pending lecture content]

