---
title: "PSYC2020L Assignment 14"
author: "[name]"
date: "2025-11-21"
output: html_document
---

## Goals for Review:

* ANOVA


## Setting Up

### 1) Rename this assignment as "Lab 14 Assignment [Last, First].Rmd" 

### 2) Check and set (if needed) your working directory. If you'd like to load packages with `library()`, please do so in this block.

```{r}

```

### 3) Download the `cake_data.csv` dataset from Canvas. Relocate it to your folder for this class and import it into R. Please display the first few rows. Additionally, please use two other functions to get a sense of the data. This data contains the breakage angle `angle` for cakes made with different `recipes`. 

```{r}

```


## Analysis of Variance (ANOVA)

### 4) You'd like to test to see if there are differences in the breakage angle for cakes made with these different recipes with a One-Way ANOVA. What are your null and alternative hypotheses for this test?



### 5) Perform this ANOVA and display the results with `summary()`.

```{r}

```

### 6) Briefly describe your results.




## Bonus Questions

These questions collectively are worth 1 extra point (10% extra) on this assignment. There is no penalty to not completing these bonus questions.

### 7) Produce means and standard deviations of `angle` for each group (`recipe`). 

```{r}

```

### 8a) Make a basic plot of the distributions of the values of `angle` by `recipe`.

```{r}

```

### 8b) Make at least three changes to this plot to make it look nice! Make sure to describe / justify each change you made (1--2 sentences or bullet points each).

```{r}

```








