---
title: "PSYC2020L Assignment 10"
author: "[name]"
date: "2025-10-24"
output: html_document
---

## Goals for Review:

* Communication


## Setting Up

### 1) Rename this assignment as "Lab 10 Assignment [Last, First].Rmd" 

### 2) Check and set (if needed) your working directory. If you'd like to load packages with `library()`, please do so in this block. Please load the `tidyverse`!. 
```{r}

```


## Communication

In this assignment, we will be working with the `msleep` dataset that comes with the `tidyverse` library. This data set has information of the sleep characteristics for different mammals! Check out the data with `?msleep` to read a little bit about the variables (feel free to comment out or remove afterwards. 

### 3) Then, use three functions to get a sense of the data.
```{r}

```


### 4a) Plot the distributions of the variables `sleep_total` and `bodywt` in two separate plots. Feel free to use whatever type of distributional geom you prefer.
```{r}

```

### 4b) Describe what you see in the plots above.



### 5a) Make a scatterplot with body weight on the x-axis and total sleep on the y-axis.
```{r}

```

### 5b) Label two points you consider interesting / outliers with the type of animal they are (Hint: first, `filter()` the data to see what the animal names are for the points you select. Then, `annotate()` the data with a `text` geom at `x` and `y` values close to the selected points. You can absolutely use another solution to label if you'd like; this is just one way).
```{r}

```

### 5c) Modify the x-axis label and y-axis label to be more readable and descriptive. Make sure to add units!
```{r}

```

### 6a) Let's say you're more interested in larger mammals. Create a new version of the dataset that only has mammals under 45kg. Also, create a new variable that represents the ratio of brain weight to body weight (Hint: divide brain weight by body weight).
```{r}

```

### 6b) Create a scatterplot that evaluates the association between the brain-to-body weight proportion variable you created (put that on the x-axis) and total hours of sleep.
```{r}

```


### 6c) Emphasize the points on the graph that represent herbivores (Hint: make a filtered version of the dataset that only includes carnivores and layer on another `geom_point()`). Add a text annotation on the plot that indicates the group is herbivores.
```{r}

```


### 7) Make one more scatterplot of two variables of interest. Make at least two annotations to the plot of things / trends you find interesting. Some ideas: label your favorite mammal on the plot with its name, add a reference line, highlight endangered animals, etc. Add a label and descriptive title to this plot. Try to make this plot as self-explanatory as possible!
```{r}

```












