---
title: "PSYC2020L Assignment 6"
author: "[name]"
date: "2025-09-26"
output: html_document
---

## Goals for Review:

* Tidying data

* Line graphs

For each question, please make sure you include code output. If the question asks for comparison or further explanation, please make sure to include that as well. 


## Setting Up

### 1) Rename this assignment as "Lab 6 Assignment [Last, First].Rmd" 

### 2) Check and set (if needed) your working directory. If you'd like to load packages with `library()`, please do so in this block. Make sure to load the `tidyverse` package!
```{r}

```


## Import

### 3) The `tidyverse` package includes a dataset called `billboard` that has song rankings for the Billboard Top 100 songs from the year 2000. Use the `head()` function to display the first six rows. Then, use the `names()` to look at the column names.
```{r}

```


## Tidy

### 4a) Is the `billboards` dataset in wide or long format? How can you tell?



### 4b) Using the `?` tool on the dataset and / or the output from \#3, please describe the contents of this dataset (specificially, an overview of what you can find in the columns) (1--2 sentences).
```{r}

```



### 4c) Given what you said in \#4a, would you want to use the `pivot_longer()` or `pivot_wider()` function on this data?



### 5) Using a pipe, let's do a few things: 

* Tidy the data as you planned to in \#4c, 

* Mutate the column with the numbered weeks so that it's just numeric (hint: the `parse_number()` function will extract numbers from a string. Feel free to review its documentation with `?`), 

* Assign this tidied version to a variable so we can continue using it.

After the pipe, print the first six rows from this tidied version.
```{r}

```


## Visualization

### 7a) Now, let's plot some of this data. Let's make a line graph that has time (week) on the x-axis and a song's ranking on the y-axis. Start by using the `ggplot()` function to make a grid with axes (no geoms yet).
```{r}

```

### 7b) Now, let's add a `geom_line()` to show the rankings of the different songs over time. Remember you will need to add a `group` argument to your aes.
```{r}

```

### 7c) Because higher is "worse" for rankings, let's reverse the y-axis so higher on the y is "better" by adding a scale layer. The function `scale_y_reverse()` will do this for us! Add it on to your plot just like you would add a new layer.
```{r}

```

### 7d) Modify the `alpha` argument within your geom to make it easier to see the patterns.
```{r}

```

### 7d) Make three more modifications to this plot.
```{r}

```

### 7e) Describe what you see in this plot (1--2 sentences).



