---
title: "PSYC2020L Assignment 2"
author: "[name]"
date: "2025-08-29"
output: html_document
---

## Goals for Review:

* Basic R functions
* Descriptive statistics

For each question, please make sure you include code output. If the question asks for comparison or further explanation, please make sure to include that as well. 


## Setting Up


### 1) Rename this assignment as "Lab 2 Assignment [Last, First].Rmd" 


### 2) Check and set your working directory
```{r}

```


## Basic R Functions


### 3) Write an R function that takes in two numbers, multiplies them together, and returns the product
```{r}

```


### 4) Write an R function that takes in three vectors, puts them in a dataframe, and returns the last two rows of the dataframe.
```{r}

```


## Data Summary

We will be using the `airquality` dataset for this assignment. Like the `iris` dataset, this comes preloaded in R. It contains information on daily air quality measurements in New York from May to September 1973.

### 5) Show me the first five rows of the `airquality` dataset
```{r}

```


### 6) Obtain the `summary()` statistics for this dataset
```{r}

```


## 7) Ozone Descriptive Statistics

### a) What is the mean?
```{r}

```

### b) What is the median?
```{r}

```


## 8) Histograms for Ozone

### a) Obtain a histogram with breaks = 10
```{r}

```

### b) Obtain a histogram with breaks = 50
```{r}

```

### c) Obtain a histogram with breaks = 200
```{r}

```

### d) Compare three histograms above and describe the differences
```{r}

```


## 9) Boxplot for Ozone

### a) Obtain a boxplot for Ozone
```{r}

```

### b) Is the data symmetrical? (Yes / No) How can you tell?

### c) Is the data skewed? (Yes / No) How can you tell?


## 10) Descriptive statistics for Temp

### a) What is the mean?
```{r}

```

### b) What is the median?
```{r}

```


## 11) Histogram for Temp

### a) Obtain a histogram with default number of breaks.
```{r}

```

## 12) Comparison between Ozone and Temp...

### a) Does the mean change?
```{r}

```

### b) Does the median change?
```{r}

```

## 13) Mode for Wind

### a) What is the mode for the Wind variable? How many times did it occur?
```{r}

```





