---
title: "PSYC2020L Assignment 2"
author: "[name]"
date: "2025-08-29"
output: html_document
---

## Goals for Review:

* Basic R functions
* Descriptive statistics

## Setting Up

### 1) Rename this assignment as "Lab 2 Assignment [Last, First].Rmd" 

### 2) Check and set your working directory

```{r}

```

## Basic R Functions

### 3) Write an R function that takes in two numbers, multiplies them together, and returns the product
```{r}

```

### 4) Write an R function that takes in three vectors, puts them in a dataframe, and returns the last two rows of the dataframe.
```{r}

```


## Data Summary

### 5) Show me the first five rows of the `airquality` dataset

```{r}

```

### 6) Obtain the `summary()` statistics for this dataset

```{r}

```
## 7) Ozone Descriptive Statistics

### a) What is the mean?
```{r}

```

### b) What is the variance?
```{r}

```

### c) What is the standard deviation?
```{r}

```

### d) What is the IQR?
```{r}

```

## 8) Histograms for Ozone

### a) Obtain a histogram with breaks = 10
```{r}

```

### b) Obtain a histogram with breaks = 50
```{r}

```

### c) Obtain a histogram with breaks = 200
```{r}

```

### d) Compare three histograms above and describe the differences
```{r}

```

## 9) Boxplot for Ozone

### a) Obtain a boxplot for Ozone
```{r}

```

### b) Is the data symmetrical? (Yes / No) How can you tell?

### c) Is the data skewed? (Yes / No) How can you tell?

## 10) Descriptive statistics for Temp

### a) What is the mean?
```{r}

```

### b) What is the variance?
```{r}

```

### c) What is the SD?
```{r}

```

### d) What is the IQR?
```{r}

```

## 11) Histogram for Temp

### a) Obtain a histogram with default number of breaks.
```{r}

```

## 12) Comparison between Ozone and Temp...
### a) Does the mean change? (Yes / No)
```{r}

```

### b) Which variable has a smaller variance? (Ozone / Temp)
```{r}

```

### c) Which variable has a smaller SD? (Ozone / Temp)
```{r}

```

### d) Which variable has a smaller IQR? (Ozone / Temp)
```{r}

```






